Imputation of Incomplete Multilevel Data with \proglang{R}

### 2023-11-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:clus}Concepts in multilevel methods}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
Concept & Details\\
\midrule
Sample unit & Units of the population from which measurements are taken in a sample.\\
Hierarchical levels & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.\\
Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.\\
Mixed effect & Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).\\
\addlinespace
ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
Stratified intercept & \\
\bottomrule
\end{tabular}
\end{table}

@@

Install non-CRAN packages if necessary:

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("amices/ggmice")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
\end{CodeInput}
\end{CodeChunk}

@@

The estimated effects in the complete data are visualized in Figure
\ref{}.

\begin{CodeChunk}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}

\begin{CodeChunk}
\begin{CodeInput}
R> # fit <- glmer(mod, family = "binomial", data = impact) # fit the model
R> # tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- select(dat, !hypox)  # remove variable
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   3.00e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.56e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
             b          df dfcom         fmi      lambda m         riv
1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
          ubar
1 1.150898e-01
2 3.244044e-02
3 2.425385e-06
4 4.674630e-03
5 4.869071e-03
6 5.059539e-03
7 1.091079e-03
8 7.950697e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10495.423 10561.192 -5238.712 10477.423     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2843  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37195                 -0.41014                  0.03052  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65802                 -1.04611                 -1.51245  
                  pupil                       ct  
                0.50405                  0.42496  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
     AIC      BIC   logLik deviance df.resid 
10500.88 10566.65 -5241.44 10482.88    11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2917  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37718                 -0.41511                  0.03067  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66935                 -1.05211                 -1.49429  
                  pupil                       ct  
                0.49013                  0.43835  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10505.026 10570.795 -5243.513 10487.026     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2908  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32339                 -0.42359                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67142                 -1.05776                 -1.51038  
                  pupil                       ct  
                0.49756                  0.42474  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.511 10585.280 -5250.755 10501.511     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2961  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33581                 -0.40871                  0.03039  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66477                 -1.05451                 -1.51858  
                  pupil                       ct  
                0.45928                  0.44419  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10522.038 10587.807 -5252.019 10504.038     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2955  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35187                 -0.40574                  0.03064  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67468                 -1.06551                 -1.52622  
                  pupil                       ct  
                0.47006                  0.44148  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

attr(,"class")
[1] "mitml.result" "list"        
\end{CodeOutput}
\begin{CodeInput}
R> # testEstimates(as.mitml.result(fit))
\end{CodeInput}
\end{CodeChunk}

@@

In this example, we demonstrate a multilevel imputation of an intercept
and slope random model with a continuous response. We use the obesity
dataset included in the \texttt{micemd}@ package, a synthetic dataset
that emulates an electronic survey in which individuals are asked to
provide information about their weight and consumption habits in
different countries. To easy the explanations we simulate data for 5
clusters and we reduce the dataset to the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `time`` Response time in minutes (inclusion-restriction variable).
\item
  `FamOb`` Family obesity history (yes or not)
\end{itemize}

In this dataset, Age and FamOb are MAR, while the weight variable is
affected by selection bias, attributed to an indirect MNAR mechanism.
This MNAR mechanism typically arises when an unobserved or omitted
variable influences both the value of the incomplete variable (in this
case, Weight) and its likelihood of being missing (denoted as R).

In the primary analysis model, BMI serves as the dependent variable,
with Age, Gender, and FamOb as predictors. Researchers, based on
observable data, are considering the inclusion of random intercept
effects, as well as introducing a random slope for the Age variable.

The main model is given by:
\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
We start by loading the data:

\begin{CodeChunk}
\begin{CodeInput}
R> #data("data_heckman", package = "micemd")
R> #dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

Now, let's begin by examining the missing patterns in the data by
cluster:

\begin{CodeChunk}
\begin{CodeInput}
R> require(patchwork)
R>   myplots <- lapply(1:5, function(i) {
+          ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
+            ggtitle(paste0("Cluster", i))+ 
+       theme(legend.position = ifelse(i==1,"bottom","none"))
+       })
R>   myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
\end{CodeInput}

@@

We observe that the missing pattern is quite similar across the
clusters. However, regarding the weight variable, we notice that is
systematically missing in cluster 3. In order to evaluate if it is
required a imputation method for 1-level or 2-level we assess the
Intraclass Correlation Coefficient (ICC) for the outcome variable, we
use the ``performance'' package:

\begin{CodeChunk}
\begin{CodeInput}
R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> performance::icc(Nulmodel)
\end{CodeInput}
\begin{CodeOutput}
# Intraclass Correlation Coefficient

    Adjusted ICC: 0.342
  Unadjusted ICC: 0.342
\end{CodeOutput}
\end{CodeChunk}

Since the ICC is above 0.1 and as the main analyis will be consider a
mixed model, we decide to use two-level (2l) imputation methods. In this
imputation process, we include all predictor variables in the main model
and since BMI is a composite variable. We also incorporate weight and
height variables.

To determine the best imputation method for these variables, we can use
the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
package, which suggests an appropriate method for MAR variables based on
the type of variable, number of observations in the cluster, and number
of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
and ``2l.2stage.norm'' for the age variable. However, after inspecting
the age density plot, we consider modifying its method to
``2l.2stage.pmm.'' For the BMI variable, we employ deterministic
imputation.

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
R> meth_mar["Age"]<-"2l.2stage.pmm" 
\end{CodeInput}
\end{CodeChunk}

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labeled as -2 and the predictor
variable labeled as 2, encompassing all variables. We need to suprime
the variable Time as this variable is not especified in the main model
and modify also in the prediction matrix the relationship between BMI,
weight and height to avoid circular predictions. Then we proced to run
the imputation model.

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
R> pred_mar[,"Cluster"]<- -2 # clustering variable
R> pred_mar[,"Time"]<-0
R> pred_mar[pred_mar==1]<-2
R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
R> ggmice::plot_pred(pred_mar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_pred-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  2.409  69.703  82.925  82.476  95.310 158.420 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-2} \end{center}

We are also contemplating the utilization of the ``pmm'' option, given
that the imputed values' weights are lower than those of the observable
values.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar_pmm,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  27.50   71.02   83.82   83.76   95.55  135.98 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mar_pmm)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-2} \end{center}

After confirming convergence, we proceed to save the results for future
use. Users are considering the possibility that the weight variable may
not have been selected randomly. It's likely that an omitted variable,
like self-esteem, could influence this selection. For instance,
individuals with lower self-esteem might have higher weight values,
impacting their willingness to provide honest information due to
embarrassment.

To address this situation, two approaches have been proposed for dealing
with Missing Not at Random (MNAR) data: pattern-mixed models and
selection models. Within pattern-mixed models, methods like the delta
method and more advanced ones like NARFS have been suggested. In
contrast, the selection model approach includes methods such as the
Heckman model, which can be particularly useful in this case. Several
methods, including Galimar2017, Hammon2021, and the recent Munoz method,
designed for two-level data, allowing for variations in intercepts and
exposure effects (random intercept and slope).

To apply the heckman method, the weight variable should be specified as
``2l.2stage.heckman'' found in the micemd package. Additionally, the
prediction matrix needs modification because this method involves
specifying two equations: one for the outcome, describing the incomplete
variable in terms of partially observed predictors (in this case, all
variables from the main model), and the other for the selection model,
explaining the probability of being observed based (R) on certain
variables.

Here for the outcome equation we consider is the same imputation model
that we used for the MAR case (main model).
\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
Regarding the selection equation, we include the same predictors as
those in the primary model, as well as a time variable. Here the time
variable serves as a restriction exclusion variable specifically
explaining the probability of being observed but not affecting the
incomplete value (Weight). In this context, we assume that the time a
user spends completing the survey serves as a proxy for the barriers
they may encounter in survey completion, such as familiarity with the
survey content or internet speed. These factors may lead the user to
skip specific questions or even the entire survey. Also we assume the
time is not have any influence on the weight assigned to the subject.
\[R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
These two equations are jointly estimated under the assumption that the
error terms are interconnected with a bivariate normal distribution. For
a more comprehensive understanding of the model and the exclusion
restriction, you can refer to (Munoz).

To integrate information from both equations, we must adjust the
prediction matrix. The cluster variable remains specified as before
(-2). In this imputation method, all the variables present in both the
selection and outcome equations are included with a random effect.
However, it's essential to distinguish which of these variables appear
in each equation.

In this framework, when a variable is shared between both equations, it
is denoted as (2). Predictors exclusive to the outcome equation are
indicated as (-4), while those exclusive to the selection equation are
labeled as (-3). Consequently, the only alteration needed in the
predictor matrix pertains to the variable ``Time.''

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mnar <- pred_mar
R> pred_mnar["Weight","Time"]<- -3
R> ggmice::plot_pred(pred_mnar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_pred-1} \end{center}

We also need to modify the method of the weight variable

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mnar <- meth_mar
R> meth_mnar["Weight"]<- "3l.2stage.heckman"
\end{CodeInput}
\end{CodeChunk}

Then we proceed to run the imputation model as before Impute the
missingness:

After executing these imputation procedures, it is essential to assess
convergence and the coherence of the imputed values. Upon examining the
weight variable, we noticed that the imputed range falls outside the
realm of plausible values (as weight should be positive). Consequently,
we are contemplating a different approach, specifically, the use of
``pmm,'' which involves imputing values from donor observations. This
approach ensures that the imputed values remain within the range of
observable values.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  10.26   73.41   86.94   88.27  101.73  186.47 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mnar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-2} \end{center}

We then run the imputation model but this time using the option of pmm,
to assure that weight values are in the range of the observable data,
this can be implemented by seeting the pmm parameter as true.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar_pmm,"long"))
\end{CodeInput}
\begin{CodeOutput}
      .imp           .id          Cluster           Time          Gender     
 Min.   : 1.0   Min.   :   1   Min.   :1.000   Min.   :1.016   Female:10720  
 1st Qu.: 3.0   1st Qu.: 528   1st Qu.:2.000   1st Qu.:3.724   Male  :10390  
 Median : 5.5   Median :1056   Median :3.000   Median :5.151                 
 Mean   : 5.5   Mean   :1056   Mean   :2.868   Mean   :5.211                 
 3rd Qu.: 8.0   3rd Qu.:1584   3rd Qu.:4.000   3rd Qu.:6.698                 
 Max.   :10.0   Max.   :2111   Max.   :5.000   Max.   :9.957                 
     Height      FamOb            Age            Weight            BMI       
 Min.   :1.450   no : 3794   Min.   :14.00   Min.   : 27.50   Min.   : 9.59  
 1st Qu.:1.642   yes:17316   1st Qu.:20.00   1st Qu.: 75.19   1st Qu.:25.84  
 Median :1.701               Median :25.00   Median : 88.88   Median :30.57  
 Mean   :1.702               Mean   :24.82   Mean   : 89.61   Mean   :31.02  
 3rd Qu.:1.760               3rd Qu.:29.00   3rd Qu.:103.78   3rd Qu.:35.89  
 Max.   :1.961               Max.   :47.00   Max.   :135.98   Max.   :63.21  
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mnar_pmm)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-2} \end{center}

Then after this modification we proceed to compare the effects on the
model, therefore we run the analysis model on each of the completed
datasets along with the dataset where the incomplete values are removed
(CC).

\begin{CodeChunk}
\begin{CodeInput}
R> library(ggplot2)
R> cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
R> plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/models-1} \end{center}

We note that there is minimal disparity in the age effect, FamObs,
across the various imputation models under consideration. Nonetheless,
with respect to Gender, we observe a notable distinction: in the M(C)AR
assumption, the effect is significant, whereas under the MNAR
assumption, it becomes insignificant. An analysis of the intercept
reveals that, under the MNAR assumption, a higher average BMI is
anticipated compared to the MAR assumption.

>>>>

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \proglang{R}}

\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with R}
\Shorttitle{Multilevel \pkg{mice}}

@@

Table 1: Concepts in multilevel methods

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels. A three-level \\
Fixed effect & Here we assume that the values of an independent variable
are fixed, i.e., \\
& the values observed in the study are representative of all values in
the \\
& in the dependent variable y e.g., blood pressure between treatments A
and B. \\
Random effect & The values of an independent variable are assumed to be
randomly drawn from \\
& admission we might select only certain hospitals that are
representative of \\
& the difference of y between individual hospitals, but rather the
variation of \\
ICC & The variability due to clustering is often measured by means of
the \\
& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
& of variance that can be attributed to the cluster-level, where a
high \\
& ICC would indicate that a lot of variability is due to the cluster \\
& structure. \\
Random effect & Multilevel models typically accommodate for variability
by including \\
& a separate group mean for each cluster. In addition to random \\
& intercepts, multilevel models can also include random coefficients \\
& and heterogeneous residual error variances across clusters {[}see
e.g. \\
& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
stratified intercept as concept..{]} \\
\end{longtable}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod <- mort ~ type + age + motor_score + pupil + ct + (1 | name) 
\end{CodeInput}
\end{CodeChunk}

@@

We will use the complete data estimates as comparative truth in this
tutorial. The estimated effects in the complete data are presented in
Table XYZ.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- select(dat, !hypox)  # remove variable
R> dat <- mutate(dat, motor_score = as.factor(motor_score))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

@@

Evaluate the convergence of the algorithm:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_trace(imp)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-conv-1} \end{center}

Analyze the imputed data:

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(
+     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
+     family = "binomial")) 
R> # tidy(pool(fit))
R> # as.mitml.result(fit)
R> # testEstimates(as.mitml.result(fit))
\end{CodeInput}
\end{CodeChunk}

The estimated effects after imputation are presented in Table XYZ.

@@

TODO: explain exclusion restriction.

Data are simulated and included in the \texttt{micemd} package. We will
use the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `time`` Response time in minutes (inclusion-restriction variable).
\end{itemize}

The imputation of these data is based on the
\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
Heckman Github repo}

Load the data:

\begin{CodeChunk}
\begin{CodeInput}
R> data("data_heckman", package = "micemd")
\end{CodeInput}
\end{CodeChunk}

Select relevant variables:

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat)
\end{CodeInput}

@@

Create predictor matrix:

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(dat)   # predictor matrix
R> pred[,"Cluster"]<- -2 # clustering variable
R> pred[, "Weight"] <- -3 #inclusion-restriction variable
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}

@@

Set the Heckman model as imputation method:

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(dat) # methods vector
R> meth["weight"]<-"2l.heckman"
\end{CodeInput}
\end{CodeChunk}

Impute the missingness:

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice(
+   dat, # dataset with missing values
+   m = 5, # number of imputations
+   maxit = 10,
+   meth = meth, #imputation method vector
+   pred = pred, #imputation predictors matrix
+   meta_method = "reml",
+   printFlag = FALSE
+ )
\end{CodeInput}
\end{CodeChunk}
